---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vpstheme

This package is in development. 

Version 1.0.0 includes useable palettes and themes for the Department of Health and the Department of Families, Fairness and Housing as well as a range of helper functions, these palettes and themes require refinement.

Version 1.0.0 also includes `scale_*_dh()` functions that enable the use of the Department of Health's palette for fills and colours in `ggplot()` objects.

In its current form, the package is useable and provides example code to enable others to extend `theme_vps_foundation()` and adapt `scale_*_dh()` for other departments.

<!-- badges: start -->

<!-- badges: end -->

The main purpose of **vpstheme** is to enable R users in the Victorian Public Service to use the colours and fonts defined in the [Brand Victoria Guidelines](https://www.vic.gov.au/brand-victoria-guidelines) to create compliant visualisations in ggplot.


## Installation

You can install the development version of **vpstheme** from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zerogetsamgow/vpstheme")
```

## Brand Victoria

At the core of **vpstheme** are elements of Brand Victoria, including colours, palettes and fonts.

### Brand Victoria colours

The Brand Victoria colour palette consists of ten primary colours, shown below.

```{r bv_primary, echo=FALSE, out.height="300px", out.width="700px"}
library(vpstheme)
scales::show_col(bv_colours$primary)
```

Incorporating the primary colour palette, the Brand Victoria secondary colour palette is designed to accommodate the great range of applications across all sectors of the brand.

The palette has been categorised in 10 colour groups, from navy to pink, but the colours do not have to be used in these sets.

Any colour can appear next to another, as long as its use follows guidelines and the colours are a harmonious colour combination.

The first colour of each secondary group is shown below.

```{r bv_secondary, echo=FALSE, out.height="500px", out.width="700px"}
scales::show_col(bv_colours$secondary)
```

A selection of other colours from each group are included in this version of the package. Each colour is named. The included colours and their names are shown below.
```{r bv_table, include = FALSE}
library(tibble)
bv.table =
    dplyr::bind_rows(
      tibble(
        name = "bv.navy",
        hex = "#201547",
        group = 1L,
        position = 1L
      ),
      tibble(
        name = "bv.royal",
        hex = "#004C97",
        group = 1L,
        position = 2L
      ),
      tibble(
        name = "bv.cobalt",
        hex = "#0072CE",
        group = 1L,
        position = 3L
      ),
      tibble(
        name = "bv.cerulean",
        hex = "#0090DA",
        group = 2L,
        position = 1L
      ),
      tibble(
        name = "bv.sky",
        hex = "#71C5E8",
        group = 2L,
        position = 3L
      ),
      tibble(
        name = "bv.teal",
        hex = "#00B2A9",
        group = 3L,
        position = 1L
      ),
      tibble(
        name = "bv.forest",
        hex = "#00573f",
        group = 4L,
        position = 1L
      ),
      tibble(
        name = "bv.mint",
        hex = "#00b140",
        group = 4L,
        position = 3L
      ),
     tibble(
        name = "bv.lime",
        hex = "#78BE20",
        group = 5L,
        position = 1L),
     tibble(
        name = "bv.sunshine",
        hex = "#F6BE00",
        group = 6L,
        position = 1L),
     tibble(
       name = "bv.orange",
       hex = "#E35205",
       group = 7L,
       position = 1L),
     tibble(
        name = "bv.amber",
        hex = "#ff9e1b",
        group = 7L,
        position = 3L),
      tibble(
        name = "bv.maroon",
        hex = "#8A2A2B",
        position = 1,
        group = 8),
      tibble(
        name = "bv.red",
        hex = "#af272f",
        position = 2,
        group = 8),
      tibble(
        name = "bv.rose",
        hex = "#D50032",
        position = 3,
        group = 8),
      tibble(
        name = "bv.purple",
        hex = "#87189D",
        position = 1,
        group = 9),
      tibble(
        name = "bv.pink",
        hex = "#C63663",
        position = 1,
        group = 10),
      tibble(
        name = "bv.blush",
        hex = "#EF4A81",
        position = 2,
        group = 10),
      tibble(
        name = "bv.musk",
        hex = "#E89CAE",
        position = 3,
        group = 10),
    )
    
```    


```{r bv_names, echo=FALSE, out.height="500px", out.width="1000px"}
library(ggplot2)
ggplot(
  data = bv.table,
  aes(x = group,
      y = -position,
      label = stringr::str_glue("{name}\n{hex}"),
      fill = name)
  ) +
  geom_point(shape = 22, size = 23, show.legend = FALSE) +
  geom_text(colour = bv.smoke, size = 3) +
  scale_fill_manual(
    values = bv_colours$complete
  ) +
  scale_y_continuous(limits = c(-3,-1), expand = c(.8,.8)) +
  scale_x_continuous(limits = c(1,10),expand = c(.05,.05)) +
  ggthemes::theme_map()

```

Each colour can be referenced directly using the {bv.\*} format or via the palette object, eg `bv_colours$primary[1]` as shown below.

```{r bv_reference, out.height="400px", out.width="700px"}
bv.navy

bv_colours$primary[2]

bv.pink

bv_colours$secondary[4]
```

### VIC font

VIC is the brand font to be used across all communications. The `.otf` files for VIC font are saved in this package, but are not currently available for use.

In place of VIC font, **vpstheme** uses [Rubik](https://fonts.google.com/specimen/Rubik) which is loaded on attachment using the **systemfonts** package.

## Departmental themes and colours

### Colour palettes

**vpstheme** utilises the Brand Victoria colours defined in the package to create Departmental palettes. Currently, the package includes palettes for:

-   [Department of Health](https://www.health.vic.gov.au/) - `dh_colours` is based on the departments standard templates, with a `primary` and `secondary` colour for each template and `tints` for each primary colour.

```{r dh_colours, out.height="400px", out.width="700px"}
scales::show_col(unlist(dh_colours), ncol = 4)

```

-   [Department of Families, Fairness and Housing](https://www.dffh.vic.gov.au/) - `dffh_colours` is based on the departments standard templates, with a `primary` and `secondary` colour for each template and `tints` for each primary colour.

```{r dffh_colours, out.height="400px", out.width="700px"}
scales::show_col(unlist(dffh_colours[1:3]), ncol = 4)

```

```{r dffh_tints, out.height="400px", out.width="700px"}
scales::show_col(unlist(dffh_colours[4:7]), ncol = 4)

```

### Themes

***vpstheme*** is designed to enable R users in the VPS to produce ***ggplot2*** products that comply with departmental style guides simply.

To enable this, ***vpstheme*** exports departmental themes that use the naming convention `theme_vps_*()`, where `*` denotes the department.

As well as utilising departmental colours these themes are rendered using the `Rubik` font.

Currently, the package includes themes for:

-   [Department of Health](https://www.health.vic.gov.au/) - `theme_vps_dh()`.

```{r theme_vps_dh, out.height="400px", out.width="700px"}
library(ggplot2)
ggplot(
  data=iris,
  aes(x=Sepal.Length, y = Petal.Length,colour=Species)) +
  geom_point(size=3) +
  scale_colour_dh(
    palette = "primary",
    labels = stringr::str_to_title) +
  scale_x_continuous(name="Sepal length")+
  scale_y_continuous(name="Petal length")+
  labs(title="Sepal and petal lengths of irises")+
  theme_vps_dh(base_colour = "white")

```

-   [Department of Families, Fairness and Housing](https://www.dffh.vic.gov.au/) - `theme_vps_dffh()`.

```{r theme_vps_dffh, out.height="400px", out.width="700px"}
ggplot(data=iris,aes(x=Sepal.Length, y = Petal.Length,colour=Species)) +
  geom_point(size=3) +
  scale_colour_manual(values=dffh_colours$core,labels=stringr::str_to_title) +
  scale_x_continuous(name="Sepal length")+
  scale_y_continuous(name="Petal length")+
  labs(title="Sepal and petal lengths of irises")+
  theme_vps_dffh()
```

The themes at present largely only differ in the background colours that can be specified. `theme_vps_dh()` has additional functionality and specifications.

-   the base_colour parameter of `theme_vps_dh` can take values `pink`, `orange`, `red` and `blue` that correspond to the department's primary colours.

-   the base_colour parameter of `theme_vps_dffh` can take values `teal`, `mint`, `blue` and `purple` that correspond to the department's core colours.

For example:

```{r theme_vps_dffh_teal, out.height="400px", out.width="700px"}
ggplot(data=iris,aes(x=Sepal.Length, y = Petal.Length,colour=Species)) +
  geom_point(size=3) +
  scale_colour_manual(values=dffh_colours$core[2:4],labels=stringr::str_to_title) +
  scale_x_continuous(name="Sepal length")+
  scale_y_continuous(name="Petal length")+
  labs(title="Sepal and petal lengths of irises")+
  theme_vps_dffh(base_colour = "teal")
```

### `theme_foundation`

Each theme is built on `theme_vps_foundation`. 

This theme is designed to be a foundation from which to build new themes, and not meant to be used directly. `theme_vps_foundation()` is a complete theme with only minimal number of elements defined. It is easier to create new themes by extending this one rather than `theme_gray()` or `theme_bw()` from ***ggplot2***, because those themes define elements deep in the hierarchy. 

This theme takes {[ggthemes]{theme_foundation}()} and sets base text family and size to `Rubik` and `15` as well as setting margins and other parameters to match departmental guidelines.

## Fill and colour functions

***vpstheme*** includes two functions that enable the use of `dh_colours` as fills and colours in `ggplot()` objects. These functions allow the user to easily use the palettes from `dh_colours` and both operate as wrappers around `scale_*_manual()`. 

`scale_colour_dh()` is used in the example above. The code for `scale_*_dh()` provide examples for other who wish to create similar functions for other departments.


## Functions

***vpstheme*** includes a small number of useful functions. These include:

-   `clean_vic_lga()` based on the [clean_state](https://runapp-aus.github.io/strayr/reference/clean_state.html) function from the ***strayr*** package this function cleans Victorian local government names into consistent formats to help with formatting and joining data.

-   `make_tints` allows you to simply create Brand Victoria compliant tint colour palettes ranging from a specified start colour to white or black in five linear increments. White/black are then excluded from the palette to leave you with four lighter/darker tints of the original colour.


## Acknowledgement

**vpstheme** was created with the **palettes** package, which provides a comprehensive library for colour vectors and colour palettes using a new family of colour classes (`palettes_colour`, and `palettes_palette`) that always print as hex codes with colour previews.

See the following vignettes to learn how to use palletes with other packages:

-   [Using palettes with ggplot2](https://mccarthy-m-g.github.io/palettes/articles/ggplot2.html)

-   [Compatibility with other colour packages](https://mccarthy-m-g.github.io/palettes/articles/compatibility.html)

